---
title: "Model Diagnostics and Curve Comparison"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model Diagnostics and Curve Comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

```{r libs}
library(mrmopt)
library(brms)
library(dplyr)
library(ggplot2)
library(purrr)
library(patchwork)
```

## Overview

After fitting a response curve model there are two questions to answer
before acting on the results:

1. **Did the sampler converge?** Bayesian inference via MCMC can fail
   silently — the sampler may return estimates even when the chains have
   not mixed properly. Diagnostics catch this.
2. **Is this the right curve shape for the data?** With six curve families
   available, it is worth checking whether the chosen type fits better or
   worse than alternatives.

This vignette covers both. It assumes you have worked through the
*Fitting and Analyzing Response Curve Models* vignette and have a fitted
model in hand.

---

## Setup: Fitting Two Curve Types

We fit both a **Gompertz** and a **Logistic** curve to the Paid Search
channel from `mrmopt_data`. This gives us a concrete comparison pair for
the second half of the vignette.

```{r fits, results = "hide"}
data(mrmopt_data)

paid_search <- mrmopt_data |> filter(channel == "Paid Search")

fit_gompertz <- fit_response(
  data   = paid_search,
  spend  = "spend",
  kpi    = "conversions",
  date   = "week",
  type   = "gompertz",
  chains = 2,
  iter   = 1000,
  warmup = 500,
  seed   = 8214
)

fit_logistic <- fit_response(
  data   = paid_search,
  spend  = "spend",
  kpi    = "conversions",
  date   = "week",
  type   = "logistic",
  chains = 2,
  iter   = 1000,
  warmup = 500,
  seed   = 3907
)
```

---

## Part 1: MCMC Diagnostics

### Why Diagnostics Matter

MCMC samplers like Stan's No-U-Turn Sampler (NUTS) explore the posterior
distribution by simulating correlated random walks across the parameter
space. When the posterior is well-behaved and the model is identifiable,
multiple independent chains will converge on the same region and mix
freely. When they do not, parameter estimates and credible intervals
cannot be trusted regardless of how plausible they look.

Nonlinear models with informative priors are generally well-behaved, but
poorly specified priors or insufficient data can produce pathological
posteriors. Running diagnostics after every fit is a good habit.

### The Diagnostics Plot

`plot(fit, type = "diagnostics")` shows two panels: trace plots for the
four parameters and a posterior predictive check.

```{r fig-diagnostics, fig.height = 7, fig.cap = "Diagnostics for the Gompertz fit. Trace plots (top) show chain mixing across b, c, d, e. The PPC (bottom) overlays observed data against 200 posterior predictive draws."}
plot(fit_gompertz, type = "diagnostics")
```

### Reading Trace Plots

Each trace plot shows the sampled parameter value at every post-warmup
iteration, coloured by chain. A well-converged model shows:

- **Good mixing**: chains interleave freely with one another — no chain
  stays consistently above or below the others.
- **Stationarity**: the traces fluctuate around a stable central value
  without drifting upward or downward over iterations.
- **Similar variance**: all chains explore roughly the same range.

Warning signs to look for:

- **Stuck chains**: one chain sits at a very different value from the
  others for extended stretches.
- **Funnel geometry**: one chain has much lower variance than the others,
  suggesting it is trapped in a narrow region of the posterior.
- **Drift or trend**: a slow upward or downward movement suggests the
  chain has not reached stationarity.

If any of these appear, try increasing `adapt_delta` (e.g.,
`control = list(adapt_delta = 0.99)`), tightening priors, or increasing
the number of warmup iterations.

### R-hat and ESS

The trace plot gives a visual impression; two numerical summaries give a
formal verdict.

**R-hat** (the potential scale reduction factor) measures convergence by
comparing within-chain variance to between-chain variance. When chains
have mixed well, these are equal and R-hat ≈ 1. Values above 1.01 are
a warning sign; above 1.05 indicate a convergence problem.

**ESS** (effective sample size) measures how many independent samples the
chains are worth, accounting for autocorrelation. Low ESS means the
posterior estimate is based on fewer independent draws than the nominal
sample count. As a rough rule of thumb, aim for ESS > 400 for reliable
estimates.

Both are extracted from the brms summary:

```{r rhat-ess}
posterior_summary <- summary(fit_gompertz)$fixed
posterior_summary[grepl("b_Intercept|c_Intercept|d_Intercept|e_Intercept",
                         rownames(posterior_summary)),
                  c("Estimate", "l-95% CI", "u-95% CI", "Rhat", "Bulk_ESS", "Tail_ESS")]
```

Healthy values have Rhat ≤ 1.01 and ESS well above 400 for both bulk
and tail. Tail ESS is particularly important for credible interval
accuracy — the 2.5th and 97.5th percentiles are estimated from fewer
effective draws than the mean.

To inspect the full brms convergence output:

```{r brms-summary, eval = FALSE}
summary(fit_gompertz)
```

### Posterior Predictive Check

The PPC panel overlays the observed KPI distribution (dark line) against
200 replicated datasets drawn from the posterior (light lines). The
question it answers is: *could the model have generated data like what
we observed?*

A well-calibrated model produces replicated datasets that span the
observed distribution. Warning signs:

- **Mode mismatch**: the observed peak falls in the tail of the
  replicated distributions — the model consistently predicts too high
  or too low.
- **Width mismatch**: replicated datasets are systematically narrower
  or wider than observed — the model is overconfident or
  underconfident about the KPI variance.
- **Skew mismatch**: replicated distributions have a different skew than
  the observed data.

For a standalone PPC with more draws:

```{r ppc-standalone, eval = FALSE}
brms::pp_check(fit_gompertz, ndraws = 400)
```

::: {.callout-note}
A good PPC does not prove the model is correct — it only shows it is
not obviously wrong. Models can pass a PPC while still being
misspecified in ways that matter for extrapolation or range analysis.
:::

---

## Part 2: Bayes R²

Bayes R² estimates the proportion of KPI variance explained by the
model, derived from posterior predictive draws rather than a single
point estimate. This gives it a natural credible interval rather than a
single number.

```{r r2}
fit_gompertz$R2
```

Interpretation guidelines:

- **High R² with tight interval**: the model fits well and the
  uncertainty about fit quality is low.
- **High R² with wide interval**: the model fits well on average but
  the credible interval reflects uncertainty — possibly because the
  posterior has high variance in the saturated region.
- **Low R²**: the model explains little variance. Check whether the
  data have sufficient spend variation and whether the curve type is
  appropriate.

R² alone should not drive curve selection — two curve types can have
similar R² while making very different predictions in unobserved spend
regions. Use R² alongside visual model comparison.

---

## Part 3: Comparing Curve Types

### When to Compare

After verifying that a model has converged, the natural next question is
whether a different curve type would fit the data better or produce more
defensible predictions. The most common comparisons are:

- **Gompertz vs. Logistic** for digital channels: the Gompertz's early
  inflection often fits performance media well; the Logistic's symmetry
  can be more appropriate when saturation is gradual.
- **Logistic vs. Log-Logistic** when unsure whether the response
  operates on a linear or proportional spend scale.
- **Gompertz vs. Reflected Gompertz** when the data suggest a slow
  early ramp rather than a quick one.

### Overlaying Response Curves

`mrm_plot_compare()` takes a named list of fitted models and overlays
their response curves, return curves, or cost-per curves on a single
axis. Names are used as legend labels.

```{r fig-compare-response, fig.height = 4, fig.cap = "Gompertz and Logistic response curves overlaid on the same axis. Both fit the observed spend range well; they diverge in the extrapolated region above ~$70k."}
mrm_plot_compare(
  models    = list(Gompertz = fit_gompertz, Logistic = fit_logistic),
  plot_type = "response"
)
```

The region where the two curves agree — broadly, within the observed
spend range — is the region where the data inform the model. The region
where they diverge is where curve choice matters most for planning and
optimization.

### Comparing Returns

The return comparison is often more revealing than the response
comparison, because it directly shows how each model characterises
the efficiency of incremental spend.

```{r fig-compare-return, fig.height = 4, fig.cap = "AR and MR curves for both models. The Gompertz's earlier inflection gives it a higher marginal return peak at lower spend; the Logistic's symmetric shape keeps MR higher for longer."}
mrm_plot_compare(
  models    = list(Gompertz = fit_gompertz, Logistic = fit_logistic),
  plot_type = "return"
)
```

Two things to look for here:

1. **Peak MR location**: does one model place peak efficiency at a
   substantially different spend level? If so, the models would give
   different investment recommendations.
2. **Decay rate**: how quickly does MR fall as spend increases? A
   steeper decay justifies tighter spend caps in optimization.

### Comparing Cost Per KPI

The cost-per comparison is the most operationally interpretable view.
It directly shows the financial cost of delivering a conversion at each
spend level.

```{r fig-compare-costper, fig.height = 4, fig.cap = "Cost per conversion curves for both models. Agreement in the observed range gives confidence that either model would support similar spend recommendations."}
mrm_plot_compare(
  models    = list(Gompertz = fit_gompertz, Logistic = fit_logistic),
  plot_type = "costper"
)
```

### Faceted Layout

When comparing more than two models or when the curves are too close to
distinguish on a single axis, use `layout = "facet"`:

```{r fig-compare-facet, fig.height = 4, fig.cap = "Faceted comparison gives each model its own panel, making the individual shapes easier to read."}
mrm_plot_compare(
  models    = list(Gompertz = fit_gompertz, Logistic = fit_logistic),
  plot_type = "response",
  layout    = "facet"
)
```

### Comparing R² Across Models

Bayes R² provides a single-number summary of relative fit quality.
When comparing multiple fitted models, collect R² values into a table:

```{r r2-compare}
list(Gompertz = fit_gompertz, Logistic = fit_logistic) |>
  map_dfr(~ as.data.frame(.x$R2), .id = "model") |>
  select(model, Estimate, Est.Error, Q2.5, Q97.5) |>
  arrange(desc(Estimate))
```

A model with meaningfully higher R² and a credible interval that does
not overlap the competitor's is a clear winner on in-sample fit. When
intervals overlap substantially, R² does not discriminate between the
models — visual comparison of curves and PPCs should carry more weight.

### Comparing Summary Ranges

The curve comparison plots reveal shape differences; the summary range
table reveals whether those differences translate into materially
different spend recommendations. If both models suggest similar
`range_min` and `range_max` values, the practical implications of
choosing one over the other are small.

```{r range-compare}
bind_rows(
  fit_gompertz$summary |> mutate(model = "Gompertz"),
  fit_logistic$summary |> mutate(model = "Logistic")
) |>
  select(model, range_min_spend, range_peak_spend, range_max_spend,
         range_min_mr, range_peak_mr, range_max_mr)
```

---

## Part 4: Comparing Across Channels

`mrm_plot_compare()` is not limited to same-channel comparisons. Passing
models from different channels produces a side-by-side view of how the
portfolio's response profiles compare — useful for identifying which
channels are most and least saturated relative to current spend levels.

The example below is `eval = FALSE` since it requires fitting all five
channels; see the *Fitting and Analyzing Response Curve Models* vignette
for the fitting loop.

```{r channel-compare, eval = FALSE}
# fits is the named list from the fitting vignette
mrm_plot_compare(fits, plot_type = "return", layout = "facet")
```

When channels operate at very different spend scales, `layout = "facet"`
with `scales = "free_y"` (the default in faceted mode) ensures each panel
uses an appropriate axis range rather than a shared one distorted by the
highest-spend channel.

---

## Decision Framework

Before moving to optimization, a model should pass each of the following
checks:

| Check | Tool | Pass criterion |
|-------|------|---------------|
| Chain convergence | Trace plots | Chains mix freely, no stuck or drifting chains |
| Numerical convergence | R-hat | All parameters Rhat ≤ 1.01 |
| Sufficient sampling | ESS | Bulk and Tail ESS > 400 for all parameters |
| Predictive calibration | PPC | Replicated distributions span observed data |
| Model fit | Bayes R² | R² meaningfully above baseline; CI not spanning 0 |
| Curve selection | `mrm_plot_compare()` | Chosen type fits as well as alternatives in observed range; sensible extrapolation |

A model that passes all six checks can be used with confidence in
downstream analysis. A model that fails the first four (MCMC-related)
checks should be re-fitted before its results are interpreted at all —
the parameter estimates themselves are unreliable.

::: {.callout-warning}
## A note on model selection and overfitting

Comparing multiple curve types on the same data and selecting the
best-fitting one introduces a mild form of model selection bias. The
selected model will tend to have higher in-sample R² than its true
out-of-sample performance. For most practical media planning purposes
this is acceptable — the curves are constrained by their parametric
form and are unlikely to severely overfit 100+ observations. But it is
worth keeping in mind when interpreting R² for a selected model.
:::
