---
title: "Fitting and Analyzing Response Curve Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting and Analyzing Response Curve Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4,
  warning = FALSE,
  message = FALSE
)
```

```{r libs}
library(mrmopt)
library(brms)
library(dplyr)
library(ggplot2)
library(patchwork)
library(purrr)
```

## Overview

This vignette walks through the full fitting and analysis workflow for a
single channel, then shows how to extend it to a multi-channel portfolio.
It assumes familiarity with the response curve theory covered in the
*Response Curve Theory and Data Requirements* vignette.

What this vignette covers:

- Preparing data and understanding the `fit_response()` API
- Reading and interpreting model output
- Visualising response curves, returns, and cost-per metrics
- Custom inference beyond observed spend levels
- Incorporating media units (e.g., impressions)
- Specifying priors when the defaults need adjustment
- Fitting a full multi-channel model list

Diagnostics (trace plots, posterior predictive checks, R-hat) and model
comparison across curve types are covered in the next vignette.

---

## The Data

`mrmopt` ships with `mrmopt_data`, a simulated dataset of five advertising
channels over two years of weekly observations.

```{r data-look}
data(mrmopt_data)
mrmopt_data
```

Each row is a channel-week observation with a spend amount and a
conversion count. For the primary walkthrough we use **Paid Search**,
which has a clean Gompertz-shaped response and no log-scale complications.

```{r data-filter}
paid_search <- mrmopt_data |>
  filter(channel == "Paid Search")
```

Before fitting it is worth looking at the raw data: how much does spend
vary, and does the scatter of spend vs. conversions suggest a saturating
shape?

```{r fig-eda, fig.cap = "Left: Paid Search weekly spend over time. Right: spend vs. conversions scatter. Spend varies meaningfully and the scatter hints at a concave relationship — both good signs for curve fitting.", fig.height = 3.5}
p_time <- ggplot(paid_search, aes(week, spend)) +
  geom_line(colour = "#2d6fa6") +
  scale_y_continuous(labels = scales::dollar_format(scale = 1e-3, suffix = "k")) +
  labs(x = NULL, y = "Weekly Spend", title = "Spend over time") +
  theme_minimal()

p_scatter <- ggplot(paid_search, aes(spend, conversions)) +
  geom_point(alpha = 0.6, colour = "#2d6fa6") +
  scale_x_continuous(labels = scales::dollar_format(scale = 1e-3, suffix = "k")) +
  labs(x = "Weekly Spend", y = "Conversions", title = "Spend vs. Conversions") +
  theme_minimal()

p_time + p_scatter
```

The scatter plot shows diminishing returns setting in — conversions level
off at higher spend levels, consistent with a Gompertz or logistic shape.

---

## A Minimal `fit_response()` Call

`fit_response()` is the entry point for model fitting. At minimum it
needs a data frame, the names of the spend and KPI columns, the date
column, and the curve type.

```{r fit-minimal, results = "hide"}
fit <- fit_response(
  data   = paid_search,
  spend  = "spend",
  kpi    = "conversions",
  date   = "week",
  type   = "gompertz",
  chains = 2,
  iter   = 1000,
  warmup = 500,
  seed   = 8214
)
```

A note on the arguments:

- **`type`**: the response curve family. See the theory vignette for
  guidance on choosing between the six options.
- **`auto = TRUE`** (the default): `mrmopt` scales the data to [0, 1]
  before fitting and sets sensible default priors automatically. This
  removes the need to specify priors by hand for most channels.
- **`anchor_strength = 0.05`** (the default via `mrmopt_prior()`): controls
  how tightly the floor parameter (`c`) is constrained near zero. The
  value is a fraction of the observed y range used as the prior SD.
  Smaller values produce narrower credible intervals at low spend. Set
  to `NULL` for loose behavior if you believe the channel generates
  baseline conversions independent of paid spend.
- **`chains`, `iter`, `warmup`**: passed directly to Stan. The values
  above are reduced for illustration speed. For production fits,
  use the defaults (`chains = 4, iter = 4000, warmup = 1000`).
- **`units`**: an optional column of media units (impressions, clicks,
  GRPs). When supplied, cost-per-unit metrics are added throughout the
  output. Covered in the [units section](#the-units-argument) below.

---

## Reading the Output

### `print()`

Calling `print(fit)` — or just typing `fit` at the console — displays a
structured summary of the fitted model.

```{r print-fit}
fit
```

The output has five sections:

1. **Header** — channel name (taken from the spend column), curve type,
   number of weeks in the training data.

2. **Current Performance** — metrics evaluated at the observed average
   weekly spend. This tells you where the channel is operating *right now*:
   - `Weekly Spend` and `KPI`: the average week.
   - `AR` (Absolute Return): conversions per dollar of total spend —
     i.e., the average efficiency of the whole budget.
   - `MR` (Marginal Return): conversions generated by the next dollar —
     i.e., the efficiency of incremental spend at the current level.
   - `CP` (Cost per conversion): dollars spent per conversion.

3. **Parameters** — the four fitted curve parameters in original data
   units. Each parameter is shown with its posterior mean and 95%
   credible interval.

4. **Response Curve Summary** — three analytically important points on
   the curve, each reported with spend, KPI, AR, MR, and CP:
   - `Range Min`: spend at peak marginal return — the most
     efficient point to begin investing. Below this level, each
     additional dollar is *more* efficient than the last.
   - `Range Peak`: spend at peak absolute return — the single most
     efficient total spend level.
   - `Range Max`: spend where marginal return has decayed to 70% of
     its peak — the point where diminishing returns accelerate
     meaningfully.
   The `%Wks` column shows what fraction of observed weeks fell below,
   within, and above this range.

5. **Bayes R²** — a posterior distribution over the proportion of
   variance in the KPI explained by the model. Higher is better; the
   credible interval reflects uncertainty about fit quality.

The underlying data for the summary is also accessible programmatically:

```{r summary-tibble}
fit$summary
```

### `mrm_params()`

For a focused view of the four curve parameters with plain-language
descriptions, use `mrm_params()`:

```{r params-summary}
mrm_params(fit)
```

Interpreting these in the context of Paid Search:

- **Ceiling (d)**: the theoretical maximum weekly conversions the
  channel could achieve with unlimited spend. The posterior uncertainty
  around this parameter is often wide — the data rarely extend into the
  fully saturated region.
- **Midpoint (e)**: the spend level at the curve's inflection point. For
  Gompertz curves this is where the curve sits at ~37% of its range, so
  the *peak efficiency* zone begins somewhat below `e`.
- **Steepness (b)**: controls how quickly the curve transitions from low
  to high response. More negative values mean a sharper transition.
- **Floor (c)**: the baseline conversion level approached at zero spend.
  With the default `anchor_strength = 0.05` this is typically close to zero.

For programmatic access to the raw posterior estimates (centre, lower,
upper credible bounds as named lists), use `mrm_params()`:

```{r params-raw}
mrm_params(fit)
```

---

## The Dashboard Plot

`mrm_plot(fit)` produces a three-panel dashboard combining the three core
views of the model.

```{r fig-dashboard, fig.height = 5, fig.cap = "The three-panel dashboard: response curve (top left), AR/MR returns (top right), and cost per conversion (bottom)."}
mrm_plot(fit)
```

**Response curve (top left)**: the fitted curve with a 95% credible band,
overlaid on the observed spend-conversion data points. The vertical
dashed lines mark the three range points from the summary. The current
average weekly spend is indicated by the solid vertical line.

**Returns (top right)**: absolute return (AR, conversions per dollar of
total spend) and marginal return (MR, conversions per next dollar) plotted
against spend. The gap between them widens as spend increases and
diminishing returns set in.

**Cost per conversion (bottom)**: the inverse of AR — dollars required
per conversion — with a credible ribbon reflecting posterior uncertainty.
Cost per conversion rises steeply as spend pushes into the saturated
region of the curve.

Two useful options:

```{r fig-dashboard-options, eval = FALSE}
# Remove range annotations for a cleaner look
mrm_plot(fit, markup = FALSE)

# When units are present, flip the x-axis to media units
mrm_plot(fit, x_var = "units")
```

---

## Individual Plot Functions

Each dashboard panel is also available as a standalone function, which
makes it easy to customise or compose them differently.

```{r fig-individual, fig.cap = "Response curve (left) and cost-per-conversion (right) composed side by side with patchwork.", fig.height = 3.5}
mrm_plot_response(fit) + mrm_plot_costper(fit)
```

The individual functions accept the same arguments as the dashboard:

```{r individual-args, eval = FALSE}
# Response curve without annotations
mrm_plot_response(fit, markup = FALSE)

# Returns using the lower credible bound of the parameters
mrm_plot_return(fit, location = "lower")

# Cost-per over a custom spend range
mrm_plot_costper(fit, xrange = c(10000, 100000))
```

---

## Custom Inference with `mrm_infer()`

`mrm_infer()` returns the full inference data frame underlying all plots.
By default it uses the observed spend range, but you can extend it to
ask counterfactual questions — for example, *what would conversions look
like if we increased spend by 50% above our current maximum?*

```{r infer-custom}
infer_extended <- mrm_infer(
  fit,
  xrange     = c(0, 100000),
  length.out = 200
)

infer_extended |>
  select(spend, center, lower, upper, ar, mr, cp) |>
  filter(spend > 70000) |>
  head(8)
```

The columns returned are:

| Column | Description |
|--------|-------------|
| `spend` | Spend value (x) |
| `center` | Posterior mean prediction |
| `lower`, `upper` | 95% credible bounds |
| `ar` | Absolute return (conversions / spend) |
| `mr` | Marginal return (Δconversions / Δspend) |
| `cp` | Cost per conversion (spend / conversions) |

For a single-value evaluation — useful in custom optimisation code or
scenario calculations — `mrm_response_function()` extracts a callable
function from the fitted model:

```{r response-function}
f <- mrm_response_function(fit)

# Predicted conversions at three hypothetical spend levels
sapply(c(30000, 50000, 80000), f)
```

---

## The `units` Argument {#the-units-argument}

When a media channel is measured in units other than dollars (impressions,
GRPs, clicks), the `units` argument adds a second perspective: how many
units does each dollar of spend buy, and what is the cost per unit of KPI?

To demonstrate, we add a simulated `impressions` column to the Paid
Search data — assuming a CPM of roughly $12 (i.e., ~83 impressions per
dollar).

```{r units-data}
set.seed(3071)
paid_search_with_units <- paid_search |>
  mutate(
    impressions = as.integer(round(spend / 12 * 1000 + rnorm(n(), 0, 500)))
  )
```

```{r fit-units, results = "hide"}
fit_units <- fit_response(
  data   = paid_search_with_units,
  spend  = "spend",
  kpi    = "conversions",
  date   = "week",
  units  = "impressions",
  type   = "gompertz",
  chains = 2,
  iter   = 1000,
  warmup = 500,
  seed   = 4417
)
```

The fitted model now carries a `cost_per_unit` (CPU) value — the average
cost per 1,000 impressions — and all inference output includes an
`impressions` column alongside `spend`.

```{r fig-units-dashboard, fig.height = 5, fig.cap = "Dashboard with impressions as the x-axis. The x-axis now shows media delivery rather than dollars, giving a units-based view of the response curve."}
mrm_plot(fit_units, x_var = "units")
```

```{r units-cpu}
# Cost per unit stored on the model
fit_units$cost_per_unit
```

When `units` are supplied, the Current Performance and Range Summary
sections of `print()` also include unit-denominated metrics.

```{r print-units}
fit_units
```

---

## Prior Specification

`auto = TRUE` works well in most cases, but there are situations where
domain knowledge should constrain the priors more tightly — for example,
when data are sparse, or when you have strong beliefs about the plausible
spend range for saturation.

### Approach 1: Automatic (default)

```{r prior-auto, eval = FALSE}
fit_auto <- fit_response(
  data = paid_search, spend = "spend", kpi = "conversions",
  date = "week", type = "gompertz"
  # auto = TRUE is the default
)
```

### Approach 2: Simplified via `mrmopt_prior()`

`mrmopt_prior()` provides scale-invariant controls on three intuitive
quantities:

- `midpoint_range`: where the curve's inflection point can fall,
  expressed as a fraction of the observed spend range. For example,
  `c(0.2, 0.7)` says the inflection point should be somewhere between
  20% and 70% of the way from minimum to maximum observed spend.
- `ceiling_max`: the ceiling parameter can be at most this multiple of
  the observed maximum KPI. A value of `2` says we don't expect the
  channel to ever deliver more than double its historical best week.
- `floor_min`: the lower asymptote should be at least this value in
  original KPI units.

```{r prior-simplified, eval = FALSE}
my_prior <- mrmopt_prior(
  midpoint_range = c(0.2, 0.7),
  ceiling_max    = 2,
  floor_min      = 0
)

fit_simplified <- fit_response(
  data  = paid_search, spend = "spend", kpi = "conversions",
  date  = "week", type = "gompertz",
  prior = my_prior
)
```

### Approach 3: Manual `brms` priors

For full control, pass a raw `brms::prior()` object. This requires
working in the scaled parameter space and is recommended only for
advanced users who need to specify distributional forms beyond the
defaults.

```{r prior-manual, eval = FALSE}
manual_prior <- c(
  brms::prior(normal(-5, 2),  nlpar = "b", ub = 0),
  brms::prior(normal(0,  2),  nlpar = "c"),
  brms::prior(normal(10, 3),  nlpar = "d"),
  brms::prior(normal(0.5, 1), nlpar = "e")
)

fit_manual <- fit_response(
  data       = paid_search, spend = "spend", kpi = "conversions",
  date       = "week", type = "gompertz",
  prior      = manual_prior,
  auto       = FALSE,
  scale_data = TRUE
)
```

::: {.callout-note}
When `auto = FALSE` and `scale_data = FALSE`, the prior must be specified
in the original data units. This is rarely necessary — scaled fitting with
`mrmopt_prior()` covers the vast majority of use cases.
:::

---

## Fitting Multiple Channels

In practice you will fit models for each channel in your media portfolio
and collect them in a named list. This is the natural input format for
model comparison (`mrms_plot_compare()`) and optimization (`opt_mix()`),
both covered in later vignettes.

The code below fits all five channels from `mrmopt_data` using their
natural curve types. Because Bayesian fitting is time-consuming, this
chunk is shown but not evaluated — run it interactively.

```{r fit-all, eval = FALSE}
channel_types <- c(
  "Paid Search"  = "gompertz",
  "Paid Social"  = "logistic",
  "Display"      = "gompertz",
  "Online Video" = "logistic",
  "TV"           = "log_logistic"
)

fits <- imap(channel_types, function(curve_type, channel_name) {
  channel_data <- mrmopt_data |>
    filter(channel == channel_name)

  fit_response(
    data   = channel_data,
    spend  = "spend",
    kpi    = "conversions",
    date   = "week",
    type   = curve_type,
    chains = 2,
    iter   = 1000,
    warmup = 500
  )
})
```

Once fitted, you can inspect any model by name:

```{r inspect-list, eval = FALSE}
fits[["Paid Social"]]
mrm_plot(fits[["TV"]])
```

Or map across the list to extract a summary metric from each:

```{r r2-table, eval = FALSE}
map_dfr(fits, ~ as.data.frame(.x$R2), .id = "channel") |>
  arrange(desc(Estimate))
```

The `fits` list is the input to `mrms_plot_compare()` in the next
vignette, where we evaluate which curve type fits each channel best and
check whether the models are well-calibrated.
