---
title: "Hierarchical Response Curves"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hierarchical Response Curves}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.5,
  message  = FALSE,
  warning  = FALSE
)
library(mrmopt)
library(dplyr)
library(ggplot2)
```

Media channels are rarely homogeneous. TV spend spans broadcast, cable, and
streaming; each partner or station within those has its own reach and response
dynamics. Fitting one curve to the whole channel discards that structure, while
fitting a separate curve to every sub-unit overfits the sparse ones.

`fit_response_hier()` fits a **single hierarchical model** for one channel where
the curve parameters are *partially pooled* across sub-channel groupings. Sparse
units borrow strength from their group; well-observed units stay close to their
own data. The degree of shrinkage is automatic and data-driven.

This vignette fits a two-level hierarchy (subtype → station), inspects the
per-level curves, visualizes the pooling, and optimizes a budget at a chosen
level of the hierarchy.

---

## Simulated data

We simulate a TV channel with three subtypes and several stations each. Curve
*shape* (growth `b`, midpoint `e`) is similar across stations; *scale* (ceiling
`d`) varies with station size. One station — `cable_b` — has only a handful of
active weeks, so we can watch partial pooling pull its estimate toward the cable
group mean.

```{r simulate}
set.seed(2026)

gompertz <- function(x, d, e, b = -3e-4) d * exp(-exp(b * (x - e)))

# subtype -> station -> (ceiling d, n weeks)
spec <- list(
  broadcast = list(bcast_a = c(d = 1500, n = 30), bcast_b = c(d = 1200, n = 30)),
  cable     = list(cable_a = c(d =  900, n = 30), cable_b = c(d = 700,  n = 6)),
  streaming = list(stream_a = c(d = 2000, n = 30))
)

make_station <- function(subtype, station, d, n) {
  spend <- runif(n, 2e3, 1.3e5)
  mu    <- gompertz(spend, d = d, e = 5e4)
  data.frame(
    week    = seq.Date(as.Date("2024-01-01"), by = "week", length.out = n),
    subtype = subtype,
    station = station,
    spend   = spend,
    aa_opps = pmax(rnorm(n, mu, 0.05 * d), 0)
  )
}

tv <- do.call(rbind, unlist(lapply(names(spec), function(st) {
  lapply(names(spec[[st]]), function(stn) {
    p <- spec[[st]][[stn]]
    make_station(st, stn, d = p[["d"]], n = p[["n"]])
  })
}), recursive = FALSE))

tv |> count(subtype, station)
```

---

## Fitting the hierarchical model

The call mirrors `fit_response()`, with one new argument: `group`, a vector of
grouping columns ordered outermost → innermost. Here `c("subtype", "station")`
fits the nested structure `(1 | subtype) + (1 | subtype:station)` on the pooled
parameters.

By default the shape parameters (`b`, `e`) and the scale parameter (`d`) are
pooled, while the floor (`c`) stays at the channel level. (Iterations are kept
modest here to keep the vignette quick.)

```{r fit, results = "hide"}
fit_tv <- fit_response_hier(
  data   = tv,
  spend  = "spend",
  kpi    = "aa_opps",
  date   = "week",
  group  = c("subtype", "station"),
  type   = "gompertz",
  chains = 2,
  iter   = 1500,
  warmup = 750
)
```

Printing the fit shows the hierarchy, the channel-level mean curve, and a
per-station table:

```{r print}
fit_tv
```

---

## Per-level summaries

`mrm_summary_hier()` returns one row per unit at every level of the hierarchy,
plus a channel-level row. The columns match the single-fit `mrm_summary()`, with
added `id` and `level` keys.

```{r summary}
mrm_summary_hier(fit_tv) |>
  select(id, level, weekly_spend, kpi_at_current, d, e) |>
  arrange(level, id)
```

Notice the ceilings (`d`) track the sizes we simulated, and the sparse station
`cable_b` is pulled toward its cable peers rather than chasing its own noisy
data.

---

## Visualizing the hierarchy

`mrm_plot_hier()` produces a **dashboard** (like `mrm_plot()` for single fits):
a response-curve panel for each level of the hierarchy, plus a shrinkage panel
for each of the shape/scale parameters (`e`, `b`, `d`).

```{r plot-dashboard, fig.height = 7}
mrm_plot_hier(fit_tv)
```

The individual panels are also exported as standalone functions when you want
just one. `mrm_plot_hier_response()` shows one curve per unit at a level,
overlaid with the channel mean (dashed); it defaults to the innermost level:

```{r plot-response}
mrm_plot_hier_response(fit_tv)
```

```{r plot-response-subtype}
mrm_plot_hier_response(fit_tv, level = "subtype")
```

`mrm_plot_hier_shrinkage()` makes partial pooling explicit. Each unit's
parameter estimate is shown with its credible interval; point size reflects the
number of observed weeks, and the dashed line marks the channel mean. Sparse
units have wider intervals and sit closer to the mean.

```{r plot-shrinkage, fig.height = 3.5}
mrm_plot_hier_shrinkage(fit_tv, param = "e")
```

### Diagnostics

`mrm_plot_hier(type = "diagnostics")` (or `mrm_plot_hier_diagnostics()`) shows
trace plots for the channel-level parameters and the partial-pooling standard
deviations, plus a posterior predictive check — the hierarchical analogue of
`mrm_plot(fit, type = "diagnostics")`.

```{r plot-diagnostics, fig.height = 7}
mrm_plot_hier(fit_tv, type = "diagnostics")
```

---

## Optimizing at any level

`as_mrmfit_list()` expands the hierarchical fit into a named list of
single-curve models — one per unit at a chosen level — that plugs directly into
`opt_mix()`. This is how you optimize "at any level of the hierarchy."

Optimize across individual stations (the innermost level):

```{r opt-stations, results = "hide"}
station_models <- as_mrmfit_list(fit_tv, level = "subtype:station")
opt_stations   <- opt_mix(station_models, budget = 400000)
```

```{r opt-stations-table}
opt_table(opt_stations)
```

Or step up a level and optimize across subtypes:

```{r opt-subtypes, results = "hide"}
subtype_models <- as_mrmfit_list(fit_tv, level = "subtype")
opt_subtypes   <- opt_mix(subtype_models, budget = 400000)
```

```{r opt-subtypes-table}
opt_table(opt_subtypes)
```

Because each unit view carries its own posterior draws, the posterior
optimization path works too and propagates pooling-aware uncertainty into the
allocation:

```{r opt-posterior, results = "hide"}
opt_post <- opt_mix(station_models, method = "posterior",
                    budget = 400000, n_draws = 100, seed = 1)
```

```{r opt-posterior-plot}
plot(opt_post, type = "posterior")
```

---

## Saturating (log-based) channels

The log-based forms (`log_logistic`, `weibull`, `reflected_weibull`) are
supported too. For these the midpoint enters as `log(e)`, which requires
`e > 0`; group-level deviations could otherwise drive it negative. The fit
handles this internally by reparameterizing the midpoint on the log scale, then
reporting `e` in the usual spend units — no change to how you call it:

```{r logform, eval = FALSE}
fit_tv_ll <- fit_response_hier(
  data  = tv,
  spend = "spend",
  kpi   = "aa_opps",
  date  = "week",
  group = c("subtype", "station"),
  type  = "log_logistic"
)
```

---

## How pooling works

- **Fixed effects** are the channel-level mean curve.
- **Random effects** at each level (subtype, then subtype:station) are drawn
  from the level above, so a station's curve is the channel mean plus its
  subtype deviation plus its own deviation.
- **Shrinkage is automatic**: units with more observations are pulled less
  toward the group mean; sparse units (like `cable_b`) are pulled more and carry
  wider credible intervals.
- **Shape pools, scale floats**: pooling concentrates on the shape parameters
  (`b`, `e`) while the ceiling (`d`) varies more freely to reflect genuine size
  differences across units.

See `?fit_response_hier` for the full argument list, including `pool` (which
parameters get group effects), `group_sd_prior`, and `min_obs`.
```
