---
title: "Media Mix Optimization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Media Mix Optimization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

```{r libs}
library(mrmopt)
library(ggplot2)
library(dplyr)
```

## Overview

Once you have fitted response curves for each media channel, `opt_mix()`
finds the budget allocation that maximises total KPI. It supports two
optimization methods:
  
- **Point estimate** (`method = "point"`): uses the posterior median
  parameters from each model. Fast — typically under one second.
- **Posterior sampling** (`method = "posterior"`): optimises over many
  posterior draws, producing a **distribution** of optimal allocations
  that reflects Bayesian uncertainty. Takes a few seconds.

This vignette walks through the full optimization workflow using the
`mrmopt_data` example dataset.

---

## 1. Fitting the Models

We start by fitting a response curve for each channel. In practice you
would compare multiple curve types per channel and select the best fit
(see the *Model Comparison & Diagnostics* vignette). Here we use
Gompertz curves for simplicity.

```{r fit-models, results = "hide"}
data(mrmopt_data)

channels <- c("Paid Search", "Paid Social", "Display",
              "Online Video", "TV")

models <- list()
for (ch in channels) {
  d <- mrmopt_data |> filter(channel == ch)
  
  # Use log_logistic for TV (log-scale channel), gompertz for the rest
  curve_type <- if (ch == "TV") "log_logistic" else "gompertz"
  
  models[[ch]] <- fit_response(
    data = d, spend = "spend", kpi = "conversions", date = "week",
    type = curve_type
  )
}
```

The result is a named list of `mrmfit` objects — the required input for
`opt_mix()`.

---

## 2. Point Estimate Optimization

The simplest call uses the default budget (total current weekly spend
across all channels) and point-estimate parameters:

```{r opt-point}
opt_point <- opt_mix(models, method = "point")
```

### Inspecting the Result

`print()` and `opt_summary()` both display a formatted console summary —
they produce the same output:

```{r print-point}
print(opt_point)
```

To retrieve a **tidy tibble** of current vs. optimal deltas for further
analysis or export, use `opt_table()`:

```{r opt-table-point}
opt_table(opt_point) |>
  select(channel, spend_delta_pct, kpi_delta_pct, cp_delta)
```

### Visualising the Allocation

`opt_plot_allocation()` shows current vs. optimal spend as grouped bars:

```{r plot-allocation, fig.cap = "Current vs. optimal spend allocation."}
opt_plot_allocation(opt_point)
```

`opt_plot_comparison()` produces a dumbbell chart showing the direction
and magnitude of spend reallocation:

```{r plot-comparison, fig.cap = "Spend reallocation from current to optimal."}
opt_plot_comparison(opt_point)
```

---

## 3. Response Curves and Returns

To understand *why* the optimizer reallocates spend, overlay the current
and optimal positions on each channel's response curve:

```{r plot-curves, fig.cap = "Response curves with current (red) and optimal (blue) spend positions.", fig.height = 6}
opt_plot_curves(opt_point)
```

`opt_plot_returns()` shows where the current and optimal points sit on
each channel's average return (AR) and marginal return (MR) curves —
this is where you can see the efficiency trade-offs driving the
allocation:

```{r plot-returns, fig.cap = "AR and MR curves at current and optimal spend.", fig.height = 6}
opt_plot_returns(opt_point)
```

---

## 4. Posterior Optimization

Point estimates ignore parameter uncertainty. Posterior optimization
runs the solver across multiple MCMC draws, producing a distribution of
solutions:

```{r opt-posterior}
opt_post <- opt_mix(
  models,
  method  = "posterior",
  n_draws = 200,
  seed    = 5291
)
```

```{r print-posterior}
print(opt_post)
```

`opt_plot_posterior()` shows the distribution of optimal spend per
channel as violin plots:

```{r plot-posterior, fig.cap = "Posterior distribution of optimal allocations across 200 draws."}
opt_plot_posterior(opt_post)
```

`opt_plot_allocation()` adds 95% credible interval error bars when used
with a posterior result:

```{r plot-alloc-post, fig.cap = "Optimal allocation with 95% CIs from posterior optimization."}
opt_plot_allocation(opt_post)
```

---

## 5. Comparing Methods

`opt_table()` returns a plain tibble, so comparing two results is a
standard join:

```{r compare}
comp <- inner_join(
  opt_table(opt_point) |>
    select(channel, spend_point = optimal_spend, kpi_point = optimal_kpi),
  opt_table(opt_post) |>
    select(channel, spend_posterior = optimal_spend, kpi_posterior = optimal_kpi),
  by = "channel"
) |>
  mutate(
    spend_diff_pct = (spend_posterior / spend_point) - 1,
    kpi_diff_pct   = (kpi_posterior   / kpi_point)   - 1
  )

comp
```

A dumbbell chart requires only a few lines of ggplot2:

```{r plot-compare, fig.cap = "Point vs. posterior optimal spend per channel."}
comp |>
  filter(channel != "TOTAL") |>
  mutate(channel = forcats::fct_reorder(channel, spend_point)) |>
  ggplot(aes(y = channel)) +
  geom_segment(aes(x = spend_point, xend = spend_posterior, yend = channel),
               linewidth = 0.8, color = "grey50") +
  geom_point(aes(x = spend_point),     color = "firebrick", size = 3) +
  geom_point(aes(x = spend_posterior), color = "steelblue", size = 3) +
  scale_x_continuous(labels = scales::dollar_format()) +
  labs(x = "Weekly Spend ($)", y = NULL,
       title = "Spend: Point vs. Posterior",
       subtitle = "Red = point  |  Blue = posterior") +
  theme_minimal()
```

---

## 6. Period Budgets

Response curves model **weekly** spend-to-KPI relationships, but
planning often involves period budgets (quarterly, annual). The
`budget` and `n_weeks` arguments translate a total period budget into
weekly optimization:

```{r opt-annual}
opt_annual <- opt_mix(
  models,
  method  = "point",
  budget  = 10000000,
  n_weeks = 52
)
print(opt_annual)
```

The solution reports both weekly and period-level totals. The
underlying assumption is that the response curve is stationary across
the period — if spend-response relationships vary seasonally, this is
an approximation.

---

## 7. Constraints

### Auto-Generated Constraints

By default, `opt_mix()` derives bounds from each model's return rate
range and a `bounds_multiplier` (default 3). This prevents the optimizer
from extrapolating far beyond observed spend levels.

### User-Supplied Constraints

For more control, pass a data frame with per-channel bounds:

```{r constraints-absolute}
my_constraints <- tibble(
  channel   = names(models),
  min_spend = c(10000, 5000, 3000, 8000, 20000),
  max_spend = c(80000, 50000, 30000, 60000, 150000)
)

opt_constrained <- opt_mix(
  models,
  method      = "point",
  budget      = 250000,
  constraints = my_constraints
)
print(opt_constrained)
```

### Share-Based Constraints

You can also set constraints as a fraction of total budget. For example,
ensuring TV gets at least 20% but no more than 40% of the budget:

```{r constraints-share}
share_constraints <- tibble(
  channel   = names(models),
  min_spend = rep(0, 5),
  max_spend = rep(200000, 5),
  min_share = c(NA, NA, NA, NA, 0.20),
  max_share = c(NA, NA, NA, NA, 0.40)
)

opt_share <- opt_mix(
  models,
  method      = "point",
  budget      = 250000,
  constraints = share_constraints
)
print(opt_share)
```

When both absolute and share-based bounds are present, the **tighter
constraint wins**.

### Fixed Channels

To lock a channel at a specific spend level (e.g., a contractual
commitment), set `fixed = TRUE`:

```{r constraints-fixed}
fixed_constraints <- tibble(
  channel   = names(models),
  min_spend = c(0, 0, 0, 0, 50000),
  max_spend = c(200000, 200000, 200000, 200000, 200000),
  fixed     = c(FALSE, FALSE, FALSE, FALSE, TRUE)
)

opt_fixed <- opt_mix(
  models,
  method      = "point",
  budget      = 250000,
  constraints = fixed_constraints
)
```

TV is locked at $50,000/week; the remaining $200,000 is optimized
across the other four channels.

```{r compare-constrained, fig.cap = "Unconstrained vs. fixed-TV optimization."}
inner_join(
  opt_table(opt_point) |> select(channel, spend_unconstrained = optimal_spend),
  opt_table(opt_fixed) |> select(channel, spend_fixed = optimal_spend),
  by = "channel"
) |>
  filter(channel != "TOTAL") |>
  mutate(channel = forcats::fct_reorder(channel, spend_unconstrained)) |>
  ggplot(aes(y = channel)) +
  geom_segment(aes(x = spend_unconstrained, xend = spend_fixed, yend = channel),
               linewidth = 0.8, color = "grey50") +
  geom_point(aes(x = spend_unconstrained), color = "firebrick", size = 3) +
  geom_point(aes(x = spend_fixed),         color = "steelblue", size = 3) +
  scale_x_continuous(labels = scales::dollar_format()) +
  labs(x = "Weekly Spend ($)", y = NULL,
       title = "Spend: Unconstrained vs. Fixed TV",
       subtitle = "Red = unconstrained  |  Blue = fixed TV") +
  theme_minimal()
```

---

## Summary

| Function / Method | Purpose |
|-------------------|---------|
| `opt_mix(method = "point")` | Fast single-solution optimization |
| `opt_mix(method = "posterior")` | Bayesian optimization with uncertainty |
| `print()` / `opt_summary()` | Formatted console summary (identical output) |
| `opt_table()` | Tidy comparison tibble with per-channel deltas |
| `opt_plot_allocation()` | Grouped bar: current vs. optimal spend or KPI |
| `opt_plot_comparison()` | Dumbbell: current → optimal spend |
| `opt_plot_curves()` | Response curves with current + optimal points |
| `opt_plot_returns()` | AR/MR curves at current + optimal |
| `opt_plot_posterior()` | Violin of posterior allocations (posterior only) |
| `inner_join(opt_table(a), opt_table(b))` | Side-by-side diff of two results |
